\documentclass[a4paper,11pt]{report}
\usepackage[margin=2.cm]{geometry}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}

\usepackage{times}
%\usepackage{txfonts}
\usepackage{bm}

\usepackage{color}

\linespread{1}
\setlength{\parskip}{1.0ex}
\setlength{\parindent}{0.0ex}

\begin{document}

\begin{center}
{\large \textsc{Introduction to Nuclear and Particle Physics}} \\[0.5cm]
\Large{Take-Home Exam II\\
Due: Monday $20^{th}$ January 2020 at 13:00}
\end{center}

{\large \textsc{Problem 1 - Neutrino Oscillation in Matter} - 30 points}

In the lecture we studied neutrino oscillation in an (effective) two-flavour system. Two neutrino flavour states, $|\nu_e\rangle$ and $|\nu_\mu\rangle$, mix with two mass states, $|\nu_1\rangle$ and $|\nu_2\rangle$, such that
\[
\begin{pmatrix}|\nu_e\rangle\\|\nu_\mu\rangle\end{pmatrix} = \begin{pmatrix}\cos\theta&\sin\theta\\-\sin\theta&\cos\theta\end{pmatrix}\cdot\begin{pmatrix}|\nu_1\rangle\\|\nu_2\rangle\end{pmatrix}
\]
For a fixed momentum $p$, the Schroedinger equations of the two mass states $|\nu_i(t)\rangle$ (with $i=1$ or $i=2$) can be written as the matrix equation
\begin{equation}\label{eq:ex1a}
i\frac{\partial}{\partial t}\begin{pmatrix}|\nu_1(t)\rangle\\|\nu_2(t)\rangle\end{pmatrix} \simeq \begin{pmatrix}p + {m_1^2}/(2p)&0\\0&p + {m_2^2}/(2p)\end{pmatrix}\cdot\begin{pmatrix}|\nu_1(t)\rangle\\|\nu_2(t)\rangle\end{pmatrix}\,,
\end{equation}
where we used the relativistic approximation $E=\sqrt{p^2+m_i^2} \simeq p+m_i^2/(2p)$ in the last step. We use natural units with $c=1$ and $\hbar=1$. Similarly, the Schroedinger equations of the two flavour states $|\nu_\alpha(t)\rangle$ (with $\alpha=e$ or $\alpha=\mu$) can be written as a matrix equation
\begin{equation}\label{eq:ex1b}
i\frac{\partial}{\partial t}\begin{pmatrix}|\nu_e(t)\rangle\\|\nu_\mu(t)\rangle\end{pmatrix} = \begin{pmatrix}H_{ee}&H_{e\mu}\\H_{\mu e}&H_{\mu\mu}\end{pmatrix}\cdot\begin{pmatrix}|\nu_e(t)\rangle\\|\nu_\mu(t)\rangle\end{pmatrix}\,.
\end{equation}

1.1) Show, that the Hamiltonian matrix of vacuum oscillations can be written as
\begin{equation}\label{eq:ex1c}
{\bf H}_{\rm vac} \equiv\begin{pmatrix}H_{ee}&H_{e\mu}\\H_{\mu e}&H_{\mu\mu}\end{pmatrix} = a\begin{pmatrix}-\cos2\theta&\sin2\theta\\\sin2\theta&\cos2\theta\end{pmatrix} + b\begin{pmatrix}1&0\\0&1\end{pmatrix}\,,
\end{equation}
and determine the energy- and mass-dependent coefficients $a$ and $b$.

{\bf Hint:} Write the flavor states $|\nu_\alpha(t)\rangle$ as a superposition of mass states $|\nu_i(t)\rangle$ and use the evolution of Eq.~(\ref{eq:ex1a}) to derive the entries in Eq.~(\ref{eq:ex1b}).

1.2) Neutrino oscillations in matter are modified due to different interactions of neutrino in matter. In particular, the interactions of electron neutrinos and electron anti-neutrinos with electrons in matter can happen via charged current exchange. Draw the individual tree-level Feynman diagrams for the processes $\nu_e+e^- \to \nu_e +e^-$ and $\overline{\nu}_e+e^- \to \overline{\nu}_e +e^-$, that are not possible for $\nu_\mu$ or $\nu_\tau$ interactions with electrons. 

1.3) These matter effects introduce an additional potential term in the Hamiltonian:
\begin{equation}\label{eq:ex1d}
{\bf H}_{\rm mat} \equiv {\bf H}_{\rm vac} + \begin{pmatrix}V&0\\0&0\end{pmatrix}\,.
\end{equation}
Show that this sum can be re-written in the form similar to Eq.~(\ref{eq:ex1c})
\begin{equation}\label{eq:ex1e}
{\bf H}_{\rm mat} = \hat{a}\begin{pmatrix}-\cos2\hat\theta&\sin2\hat\theta\\\sin2\hat\theta&\cos2\hat\theta\end{pmatrix} + \hat{b}\begin{pmatrix}1&0\\0&1\end{pmatrix}\,,
\end{equation}
and give $\hat{a}$, $\hat{b}$ and $\sin2\hat{\theta}$ as a function of $a$, $b$, $\theta$ and $V_W$.

{\bf Hint:} Use the Eqs.~(\ref{eq:ex1c}), (\ref{eq:ex1d}) and (\ref{eq:ex1e}).

1.4) The neutrino flavour oscillation probability in matter looks very similar to that in vacuum. The oscillation amplitude is proportional to the new effective mixing angle as:
\begin{equation}
P_{\nu_\mu\to\nu_e} = \sin^22\hat{\theta}\sin^2\left(\frac{\hat{a}}{a}\frac{\Delta m^2\ell}{4p}\right)\,.
\end{equation}
Use the results in question 1.3 and show that the oscillation amplitude $\sin^22\hat{\theta}$ becomes maximal at a neutrino momentum corresponding to 
\begin{equation}
2a\cos2\theta = V\,.
\end{equation} 
This is the so-called Mikheyev-Smirnov-Wolfenstein (MSW) resonance.
\vspace{1cm}

{\large \textsc{Problem 2 - Helicity and Chirality} - 15 points}

A left-handed electron has a total energy $E$ and momentum ${\bf p}$ in the laboratory system. Its state $|u_\downarrow\rangle$ can be described as a superposition of left- and right-handed chirality states, $|u_\downarrow^\infty\rangle$ and $|u_\uparrow^\infty\rangle$, as 
\begin{equation}\label{eq:ex2a}
|u_\downarrow\rangle =\frac{1}{2}\sqrt{\frac{E+m}{E}}\left[ \left(1+\frac{|{\bf p}|}{E+m}\right)|u_\downarrow^\infty\rangle+\left(1-\frac{|{\bf p}|}{E+m}\right)|u_\uparrow^\infty\rangle\right]\,,
\end{equation}
where $m$ is the mass of the electron and we use natural units with $c=1$ and $\hbar=1$.

2.1) We want to consider the relativistic limit $m\ll E$. Show, that to leading order in $m/E$ the left-handed helicity state is given
\[
|u_\downarrow\rangle \simeq |u_\downarrow^\infty\rangle +\frac{m}{2E}|u_\uparrow^\infty\rangle
\]
%{\bf Hint:} Taylor-expand with respect to $m/E$.

2.2) Consider now the state of Eq.~(\ref{eq:ex2a}) in the rest-frame of the electron. What is the states composition in terms of left- and right-handed chirality states? What are the probabilities of finding the state in the left- or right-handed chirality state?

2.3) Show that in the case $m=0$, the left-handed helicity state becomes identical to the left-handed chirality state.
%2.2) Consider a Lorentz boost in the direction of the electron's momentum. The momentum of the electron in the new boosted reference system with observer velocity $v$ (measured in the original laboratory frame) is given in natural units as
%\begin{equation}
%p' = \gamma p - \gamma\beta E\,,
%\end{equation}  
%where $E$ is the electron's energy, $\beta = v/c$ and $\gamma = 1/\sqrt{1-\beta^2}$. Determine the critical velocity $v$ that is necessary to flip the helicity of the electron. If the electron had no mass, would it be possible to find an observer frame where the left-handed electron becomes a right-handed electron? Explain.

\vspace{1cm}

{\large \textsc{Problem 3 - Neutrino Scattering} - 15 points}

Consider the scattering processes $\nu_\mu +e^- \to \nu_\mu +e^-$. 

3.1) Draw the leading order tree-level diagram.

3.2) Calculate the ratio of the cross sections
\begin{equation}
\frac{\sigma(\nu_\mu +e_L^- \to \nu_\mu +e_L^-)}{\sigma(\nu_\mu +e_R^- \to \nu_\mu +e_R^-)}\,,
\end{equation} 
where the index $e^-_L$ is a left-handed and $e^-_R$ is a right-handed helicity electrons. (The neutrino is always assumed to be left-handed and we don't show the index there. Similarly, the anti-neutrino is always right-handed.)

{\bf Hint:} Start by considering the coupling of weak interactions to left-handed and right-handed leptons in the Standard Model. What are the relevant ``charges'' in the weak interactions for neutrinos and electrons?

3.3) Show that 
%\begin{equation}
%\frac{\sigma(\nu_\mu +e_L^- \to \nu_\mu +e_L^-)}{\sigma(\nu_\mu +\mu_L^- \to \nu_\mu +\mu_L^-)}\,.
%\end{equation} 
\begin{equation}
\frac{\sigma(\nu_\mu +e_L^- \to \nu_\mu +e_L^-)}{\sigma(\overline\nu_\mu +e_R^+ \to \overline\nu_\mu +e_R^+)}=1\,,
\end{equation}
to all orders in scattering theory.

{\bf Hint:} Consider the invariance of the Standard Model under certain combinations of discrete transformations.

\vspace{1cm}

{\large \textsc{Problem 4 - PET Scanning} - 25 points}

Fluorine-18 ($^{18}_{\;\; 9}$F) is one of the early tracers used in positron emission tomography (PET), having been in use since the 1960s. It is produced by bombarding  $^{18}_{\;\; 8}$O nuclei with high energy protons (typically ~18 MeV protons). In the process the $^{18}_{\;\; 9}$F is created together with another particle. The $^{18}_{\;\; 9}$F nucleus is unstable and decays via $\beta^+$.

4.1) Indicate the complete reaction for the production and decay processes.

The positron from the decay will annihilate with an electron in the tissue and produce to back-to-back photons.

4.2) Draw the Feynman diagram for the decay process.

4.3) Determine the Q-value of the decay in MeV.

4.4) Determine the half-time, T$_{1/2}$, for the decay using Sargent's rule with $|M_{fi}|\approx 1$ and $\theta_C = 13$ degrees.

The position of the annihilation is determined by measuring the arrival of the photons at the detectors. 

4.5) If the photons are detected in 2 detectors with a time difference of 0.42ns, how far from the midpoint between the detectors did the annihilation happen?

\vspace{1cm}

{\large \textsc{Problem 5 - Heavy Ion collisions} - 50 points}

The international particle physics community is considering building a collider with a circumference of 100 km at CERN. It is called the FCC, which stands for the Future Circular Collider. The machine, if built, would be able to collide protons (p+p collisions) and heavy-ions f.ex. $^{238}$U + $^{238}$U for studies of the Quark Gluon Plasma. In case the case of p+p collisions the total collision energy would be $\sqrt{s}$ = 100 TeV.

5.1) Calculate the energy of a $^{238}$U beam in the FCC.

5.2) List the relevant quantum numbers for the ground state of $^{238}$U. Calculate the volume energy, Coulomb energy, surface energy, and symmetry energy of $^{238}$U. Calculate the 1- neutron and 1-proton separation energies from $^{238}$U.

5.3) What is the shape of $^{238}$U in its ground state? Calculate the average radius of $^{238}$U. What is the length of the longest major axis? Calculate the geometrical cross section of the $^{238}$U nucleus expressed in units of barns.
 
5.4) Write up the electrostatic potential (in MeV units) between an elementary charge and a $^{238}$U nucleus as a function of distance (in fm). Calculate at which value of the impact parameter b (in fm) between two passing $^{238}$U nuclei the electric field can no longer eject a proton from one of the U nuclei.

5.5) Use the previously obtained distance to estimate the magnitude of the total ’electromagnetic dissociation’ cross section (the probability for the nucleus to break up due to Coulomb forces).

5.6) Calculate the Q-value for emission of an $\alpha$ particle from $^{238}$U. Calculate the Q-value for $\beta^-$ emission from $^{238}$U. What do you conclude?

5.7) Calculate the Gamow decay factor for α-particle decay from $^{238}$U and compare it to the similar factor for $^{208}$Pb.

\vspace{1cm}

{\large \textsc{Problem 6 - Fission reactor} - 15 points}

In a traditional fission reactor, the fission of $^{235}$U is induced by a neutron and the fission fragments are $^{92}$Rb and $^{140}$Cs.

6.1) Determine the energy released (in MeV) per fission. (Ignore the (negligible) contributions from the pairing term).

The reaction is used to power a 100 MW nuclear reactor whose core is a sphere of radius 0.75 m. 

6.2) If an average of 1 neutron per fission escapes the core, what is the neutron flux at the outer surface of the reactor in m$^{-2}$s$^{−1}$?

The core is surrounded by 1.5 m$^3$ of ideal gas maintained at a pressure of $1\times 10^5$ Pa and a temperature of 300 K. All neutrons escaping the reactor core pass through the gas.

6.3) If the interaction cross-section between the neutrons and the gas is 1 mb, what is the rate of neutron interactions in the gas?

\end{document}